---
title: "02_trait_hab_ordination"
output: html_notebook
---

This script tries to visualise the similarity or difference of traits in different habitats.  


# Packages
```{r}
library(tidyverse)
library(vegan)
```

# load data
```{r}
#trait.plot.clim.data <- read_rds("../data/trait.plot.clim.data.rds")
library(readr)
trait.plot.data <- read_csv("../data/Trait.means.site.plot.year.csv")
```



# Prepare the data


### add unique id
```{r}
data <- trait.plot.data
names(data)
table(duplicated(data$obs.ID))
```


### make it tidy

```{r}
data.all.t <- data %>% 
  select(SITEABBR, obs.ID, TARGET_HAB, light.mean.wt, moist.mean.wt, ph.mean.wt, fert.mean.wt, comp.mean.wt, stress.mean.wt, rud.mean.wt) %>% 
  gather(key = Index, value = ellval, light.mean.wt:rud.mean.wt)

#simplify index names
data.all.t$Index <- recode_factor(data.all.t$Index,
                                  light.mean.wt = "light", 
                                  moist.mean.wt = "moist",
                                  ph.mean.wt = "pH", 
                                  fert.mean.wt = "fert", 
                                  comp.mean.wt = "comp",
                                  stress.mean.wt = "stress", 
                                  rud.mean.wt = "rud") 

data.all.t
```



# Visualise


### Radar chart

prepare data

```{r}
#filter the data
unique(data.all.t$TARGET_HAB)
#the filter function is from dplyr, part of the tidyverse
chart.data <- filter(data.all.t, 
                     TARGET_HAB %in% c("Upland blanket bog", 
                                       "Lowland raised bog", 
                                       "Lowland fen"))
chart.data
```




```{r}
#define radar coordinate system 
coord_radar <- function (theta = "x", start = 0, direction = 1) 
{
  theta <- match.arg(theta, c("x", "y"))
  r <- if (theta == "x") 
    "y"
  else "x"
  ggproto("CordRadar", CoordPolar, theta = theta, r = r, start = start, 
          direction = sign(direction),
          is_linear = function(coord) TRUE)
}
```


```{r}
ggplot(chart.data, aes(x = Index, y = ellval)) +
  geom_polygon(aes(group = obs.ID, color = SITEABBR), 
             fill = NA, size = 0.01, alpha = 0.1) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  xlab("") + ylab("") +
  guides(color = guide_legend(ncol=2)) +
  coord_radar()
```



### Radar graph compare a few habitats



prepare data

```{r}
#filter the data
chart.data <- filter(data.all.t, TARGET_HAB %in% c("Upland blanket bog", "Lowland raised bog", "Lowland fen"))
```

```{r}
ggplot(chart.data, aes(x = Index, y = ellval)) +
  geom_polygon(aes(group = obs.ID, color = SITEABBR), 
             fill = NA, size = 0.01, alpha = 0.1) +
  facet_wrap(~TARGET_HAB) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  xlab("") + ylab("") +
  guides(color = guide_legend(ncol=2)) +
  coord_radar()
```

### Radar chart mkultiple

```{r}
#filter the data
chart.data <- data.all.t
```


```{r}
ggplot(chart.data, aes(x = Index, y = ellval)) +
  geom_polygon(aes(group = obs.ID, color = TARGET_HAB), 
               fill = NA, size = 0.5, alpha = 1) +
  facet_wrap(~ SITEABBR) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  xlab("") + ylab("") +
  guides(color = guide_legend(ncol=2)) +
  coord_radar()
```



### Radar chart for all data by habitat

```{r}
#the filter function is from dplyr, part of the tidyverse
data.all <- trait.plot.clim.data
```
### add unique id
```{r}
data.all$obs.ID <- paste(data.all$SITEABBR, data.all$PLOT_ID, sep = "_")
```


### make it tidy

```{r}
data.all.t <- data.all %>% 
  select(SITEABBR, obs.ID, TARGET_HAB, TARGET_HAB, light.mean.wt, 
         moist.mean.wt, ph.mean.wt, fert.mean.wt) %>% 
  gather(key = Index, value = ellval, light.mean.wt:fert.mean.wt) 
data.all.t

data.all.t$Index <- recode_factor(data.all.t$Index,
                                  light.mean.wt = "light", 
                                  moist.mean.wt = "moist",
                                  ph.mean.wt = "pH", 
                                  fert.mean.wt = "fert") 

```



```{r}
#define radar coordinate system 
coord_radar <- function (theta = "x", start = 0, direction = 1) 
{
  theta <- match.arg(theta, c("x", "y"))
  r <- if (theta == "x") 
    "y"
  else "x"
  ggproto("CordRadar", CoordPolar, theta = theta, r = r, start = start, 
          direction = sign(direction),
          is_linear = function(coord) TRUE)
}
```


```{r}
ggplot(data.all.t, aes(x = Index, y = ellval)) +
  geom_polygon(aes(group = obs.ID, color = TARGET_HAB), 
               fill = NA, size = 0.5, alpha = 1,
               show.legend = FALSE) +
  geom_line(aes(group = obs.ID, color = TARGET_HAB), size = 0.5, alpha = 1,
            show.legend = FALSE) +
  facet_wrap(~ TARGET_HAB) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  xlab("") + ylab("") +
  #guides(color = guide_legend(ncol=2)) +
  coord_radar()
```


try with log scale

```{r}
ggplot(data.all.t, aes(x = Index, y = ellval)) +
  geom_polygon(aes(group = obs.ID, color = TARGET_HAB), 
               fill = NA, size = 0.5, alpha = 1,
               show.legend = FALSE) +
  geom_line(aes(group = obs.ID, color = TARGET_HAB), size = 0.5, alpha = 1,
            show.legend = FALSE) +
  #scale_y_log10() +
  scale_y_sqrt() +
  facet_wrap( ~ TARGET_HAB) +
  theme(strip.text.x = element_text(size = rel(0.8)),
        axis.text.x = element_text(size = rel(0.8)),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  xlab("") + ylab("") +
  guides(color = guide_legend(ncol=2)) +
  coord_radar()
```






### extract trait data

we're going to use the weighted trait data.  Lets remind ourselves of the column headings: 

```{r}
names(data)
```


```{r}
#the select function is from dplyr, part of the tidyverse
traits <- select(data, light.mean.wt:fert.mean.wt)
traits
```

### extract plot data

```{r}
plots <- select(data, SITEABBR, SITE_HAB,  SITECODE, PLOT_ID, TARGET_HAB, TARGET_HAB)
plots
```

### extract environmental data

```{r}
env <- select(data, Tmax:mean.sum.precip)
env
```

# Ordinate trait data

## Nonmetric Multidimensional Scaling

```{r}
library(vegan)
set.seed(201) # this allows us to reproduce the same result in the future
meta.nmds <- metaMDS(traits) #no transformation of species data is made here prior to bray curtis dissimilarities being calculated. (Bray Curtis is the default in R).
#str(meta.nmds) # gives stress value for plot
#stressplot(meta.nmds) # To gain the stress plot for stress values for your MDS
```

### Merge ordination scores with plot info

```{r}
plots.nmds <- plots #make a copy of our plot data

plots.nmds$NMDS1<-meta.nmds$points[ ,1] #this puts the NMDS scores for the plots into a new dataframe. you could put them into an existing one if you preferred.
plots.nmds$NMDS2<-meta.nmds$points[ ,2] 
```


```{r}
##species data
stems<-colSums(traits) #global sum of indices for each ellenberg value
trts <- data.frame(scores(meta.nmds, display = "species")) #dataframe of trait scores for plotting
trts$species <- row.names(trts) # making a column with trait names
trts$colsums <- stems #adding the colSums from above
trts<-trts[!is.na(trts$NMDS1) & !is.na(trts$NMDS2),] #removes NAs
trts.colmedian <- median(trts$colsums) #create an object that is the median of the abundance of the measured species
trts.colmean <- mean(trts$colsums) #creates a mean instead if you wish to use
trts2 <- subset(trts,trts$colsums > trts.colmean) #select the most abundant species. Could discard fewer by going something like - trts$colsums>(trts.colmedian/2) instead
trts2$species <- factor(trts2$species) #otherwise factor doesn't drop unused levels and it will throw an error
```


```{r}
ggplot(data = plots.nmds, aes(y = NMDS2, x = NMDS1)) + #sets up the plot
         geom_point(aes(shape = SITEABBR), size = 3) +
   
```





### Detrended correspondence analysis (DECORANA)

```{r}
library(vegan)
dca.1 <- decorana(traits) #carry out ordination
dca.1
plot(dca.1)
```

### Merge ordination scores with plot info

```{r}
plots.dca <- plots #make a copy of our plot data

plots.dca$DCA1 <- dca.1$

```











```{r}
ordiplot(ord = dca.1, display = "sites", type = "n")
points(dca.1, display = "sites",  = plots$SITEABBR)

#legend("topright", legend = levels(droplevels(plots$SITEABBR)), col = plots$SITEABBR)
```



```{r}
ordiplot(dca.1, type="n", xlim = c(-4,9), ylim = c(-4,5)) #blank plot
title(main="all LTMN sites ordination on spp freq DCA")

colvec <- rainbow(length(levels(droplevels(plot.mat$TARGET_HAB)))) #colour vector, random colours
pchvec <- c(1:length(levels(droplevels(plot.mat$TARGET_HAB)))) #symbol vector (gets tricky above 21)

#with(plot.mat, levels(droplevels(plot.mat$TARGET_HAB))) #note 'droplevels' to get rid of unused levels

with(plot.mat, points(dca.1, display = "sites", col = colvec[droplevels(plot.mat$TARGET_HAB)], pch = 20, cex = 0.8))

with(plot.mat, legend("topright", legend = levels(droplevels(plot.mat$TARGET_HAB)), bty = "n",
                      col = colvec, pch = 20, pt.bg = colvec, cex = 0.6, pt.cex = 1))



```



```{r}
dune.nmds.gg1 <- ggplot(data = dca.1, aes(y = DCA1, x = DCA2)) 
                        
                        
                        + #sets up the plot. brackets around the entire thing to make it draw automatically
   geom_path(data = df_ell.dune.management, aes(x = NMDS1, y = NMDS2, group = Management, alpha=Management))+ #this is the ellipse, seperate ones by Site. If you didn't change the "alpha" (the shade) then you need to keep the "group 
   scale_alpha_manual(guide = FALSE,values=c(0.3, 0.5, 0.7, 0.9))+ #sets the shade for the ellipse
   geom_point(aes(shape = Management), size = 3) + #puts the site points in from the ordination, shape determined by site, size refers to size of point
  # geom_text(data=trts2, aes(x=trts2$NMDS1, y=trts2$NMDS2, label=species), size = 3.3, hjust=1.1)+ #labelling the species. hjust used to shift them slightly from their points
      annotate("text",x = NMDS.mean$NMDS1,y = NMDS.mean$NMDS2,label=NMDS.mean$group) + #labels for the centroids - I haven't used this since we have a legend. but you could also dithc the legend, but plot will get v messy
       geom_segment(data = env.scores.dune,
                    aes(x = 0, xend = mult*NMDS1, y = 0, yend = mult*NMDS2),
                    arrow = arrow(length = unit(0.25, "cm")), colour = "grey") + #arrows for envfit.  doubled the length for similarity to the plot() function. NB check ?envfit regarding arrow length if not familiar with lengths
       geom_text(data = env.scores.dune, #labels the environmental variable arrows * "mult" as for the arrows
                 aes(x = mult*NMDS1, y = mult*NMDS2, label=env.variables),
                 size = 5,
                 hjust = -0.5)+
   #geom_point(data=trts2, alpha = .6, shape = 4)+ #these are the species points, made lighter and a specific shape
   scale_shape_manual(values = c(1,8,19,5))+ #sets the shape of the plot points instead of using whatever ggplot2 automatically provides
   coord_cartesian(xlim = c(-1,1.5))+  ## NB this changes the visible area of the plot only (this is a good thing, apparently). Can also specify ylim. Here in case you want to set xaxis manually.
   theme_bw())
```



```{r}


mult <- 2 #multiplier for the arrows and text for envfit below. You can change this and then rerun the plot command.
(dune.nmds.gg1 <- ggplot(data = dune.NMDS.data, aes(y = NMDS2, x = NMDS1))+ #sets up the plot. brackets around the entire thing to make it draw automatically
   geom_path(data = df_ell.dune.management, aes(x = NMDS1, y = NMDS2, group = Management, alpha=Management))+ #this is the ellipse, seperate ones by Site. If you didn't change the "alpha" (the shade) then you need to keep the "group 
   scale_alpha_manual(guide = FALSE,values=c(0.3, 0.5, 0.7, 0.9))+ #sets the shade for the ellipse
   geom_point(aes(shape = Management), size = 3) + #puts the site points in from the ordination, shape determined by site, size refers to size of point
  # geom_text(data=trts2, aes(x=trts2$NMDS1, y=trts2$NMDS2, label=species), size = 3.3, hjust=1.1)+ #labelling the species. hjust used to shift them slightly from their points
      annotate("text",x = NMDS.mean$NMDS1,y = NMDS.mean$NMDS2,label=NMDS.mean$group) + #labels for the centroids - I haven't used this since we have a legend. but you could also dithc the legend, but plot will get v messy
       geom_segment(data = env.scores.dune,
                    aes(x = 0, xend = mult*NMDS1, y = 0, yend = mult*NMDS2),
                    arrow = arrow(length = unit(0.25, "cm")), colour = "grey") + #arrows for envfit.  doubled the length for similarity to the plot() function. NB check ?envfit regarding arrow length if not familiar with lengths
       geom_text(data = env.scores.dune, #labels the environmental variable arrows * "mult" as for the arrows
                 aes(x = mult*NMDS1, y = mult*NMDS2, label=env.variables),
                 size = 5,
                 hjust = -0.5)+
   #geom_point(data=trts2, alpha = .6, shape = 4)+ #these are the species points, made lighter and a specific shape
   scale_shape_manual(values = c(1,8,19,5))+ #sets the shape of the plot points instead of using whatever ggplot2 automatically provides
   coord_cartesian(xlim = c(-1,1.5))+  ## NB this changes the visible area of the plot only (this is a good thing, apparently). Can also specify ylim. Here in case you want to set xaxis manually.
   theme_bw())
```
















